---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# orbital

<!-- badges: start -->
<!-- badges: end -->

The goal of orbital is to enable running predictions of tidymodels [workflows](https://workflows.tidymodels.org/) inside databases.

## Installation

To install it, use:

``` r
install.packages("orbital")
```

You can install the development version of orbital from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tidymodels/orbital")
```

## Example

Given a fitted workflow

```{r}
library(tidymodels)

rec_spec <- recipe(mpg ~ ., data = mtcars) %>%
  step_normalize(all_numeric_predictors())

lm_spec <- linear_reg()

wf_spec <- workflow(rec_spec, lm_spec)

wf_fit <- fit(wf_spec, mtcars)
```

You can predict with it like normal.

```{r}
predict(wf_fit, mtcars)
```

We can get the same results by first creating an orbital object

```{r}
library(orbital)
orbital_obj <- orbital(wf_fit)
orbital_obj
```

and then "predicting" with it using `orbital_predict()` to get the same results

```{r}
mtcars %>%
  as_tibble() %>%
  orbital_predict(orbital_obj)
```