---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# orbital

<!-- badges: start -->
[![R-CMD-check](https://github.com/tidymodels/orbital/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tidymodels/orbital/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/orbital)](https://CRAN.R-project.org/package=orbital)
[![Codecov test coverage](https://codecov.io/gh/tidymodels/orbital/branch/main/graph/badge.svg)](https://app.codecov.io/gh/tidymodels/orbital?branch=main)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of orbital is to enable running predictions of tidymodels [workflows](https://workflows.tidymodels.org/) inside databases.

## Installation

To install it, use:

``` r
install.packages("orbital")
```

You can install the development version of orbital from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tidymodels/orbital")
```

## Example

Given a fitted workflow

```{r}
#| message: false
library(tidymodels)

rec_spec <- recipe(mpg ~ ., data = mtcars) |>
  step_normalize(all_numeric_predictors())

lm_spec <- linear_reg()

wf_spec <- workflow(rec_spec, lm_spec)

wf_fit <- fit(wf_spec, mtcars)
```

You can predict with it like normal.

```{r}
predict(wf_fit, mtcars)
```

We can get the same results by first creating an orbital object

```{r}
library(orbital)
orbital_obj <- orbital(wf_fit)
orbital_obj
```

and then "predicting" with it using `predict()` to get the same results

```{r}
predict(orbital_obj, as_tibble(mtcars))
```

you can also predict in most SQL databases

```{r}
library(DBI)
library(RSQLite)

con <- dbConnect(SQLite(), path = ":memory:")
db_mtcars <- copy_to(con, mtcars)

predict(orbital_obj, db_mtcars)
```

and spark databases

```{r}
library(sparklyr)
sc <- spark_connect(master = "local")

sc_mtcars <- copy_to(sc, mtcars, overwrite = TRUE)

predict(orbital_obj, sc_mtcars)
```

# Supported models and recipes steps

Full list of supported models and recipes steps can be found here: `vignette("supported-models")`.

## contributing

This project is released with a [Contributor Code of Conduct](https://github.com/tidymodels/orbital/blob/main/.github/CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

- For questions and discussions about tidymodels packages, modeling, and machine learning, please [post on Posit Community](https://forum.posit.co/new-topic?category_id=15&tags=tidymodels,question).

- If you think you have encountered a bug, please [submit an issue](https://github.com/tidymodels/orbital/issues).

- Either way, learn how to create and share a [reprex](https://reprex.tidyverse.org/articles/articles/learn-reprex.html) (a minimal, reproducible example), to clearly communicate about your code.

- Check out further details on [contributing guidelines for tidymodels packages](https://www.tidymodels.org/contribute/) and [how to get help](https://www.tidymodels.org/help/).
